Cluster Difference Imaging Photometric Survey (CDIPS) II: A Jupiter-Sized Planet Candidate in IC~2602

### 2020-06-02 paper/ms.tex
update paper

\begin{abstract}
We report analysis and arguments for the planetary nature of the
  transiting planet TOI 837.01.
We characterize the system using data from the NASA {\it Transiting
  Exoplanet Survey Satellite} (TESS), the ESA Gaia mission,
  ground-based photometry, and spectroscopy from CTIO1.5/CHIRON,
  MPG2.2/FEROS, and AAT/Veloce.
We find that TOI$\,$837 is a $T=9.9$ G-dwarf in IC~2602 (the
``Southern Pleiades''). 
The star is therefore 40$\pm$10 million years old.
The validated planet is warm ($P = 8.YY\,{\rm d}$) and likely
Jupiter-sized. 
However, its transits are grazing: $b > 0.92$ at 3$\sigma$.
From transit photometry alone, the planetary size lies within
0.58--3.87 $R_{\rm Jup}$, due to the degeneracy
between the planetary size and its impact parameter.
From radial velocity monitoring, we find 
that its mass is less than 2.XX $M_{\rm Jup}$ (3$\sigma$).
This mass limit requires the planetary size to be below $\approx$2.X$R_{\rm Jup}$.
Non-planetary explanations for the data, most notably
hierarchical or background eclipsing binaries, are argued to be
highly unlikely on statistical grounds.
TOI 837.01 therefore tentatively joins the ranks of adolescent exoplanets orbiting
bright host stars. Further analysis of its mass, atmosphere, and obliquity could
confirm its planetary nature beyond doubt, and are worthwhile because
young planets do interesting things.
\end{abstract}

@@

\subsection{Ground-based Time-Series Photometry}

@@

\tn\ has been reported as a member of \cn\ by multiple independent 
investigators \citep{Kharchenko_et_al_2013,cantatgaudin_gaia_2018}.
%FIXME FIXME finish these citations
%include: Damiani+2019, Cantat-Gaudin+2020 who explicitly listed as
%IC2602

@@

@@

\subsection{Validation of \pn}
\label{subsec:validation}

@@

Figure~\ref{fig:fpscenario} provides a visual summary of
the possible astrophysical false positive scenarios.
If the \tn\ signal is not planetary, it must come from
the near-total eclipse of a background star (a BEB), or else a hierarchical companion
of the primary star (an HEB).
It must also be verified that to the angular
resolution possible, the transit signal originates from
\tn, and not a neighboring star (an NEB).

For these scenarios to be viable, the flux from \tn\ and the true
eclipsing binary host would blend together, reducing the ``true''
eclipse depth $\delta_{\rm true}$ to produce the observed depth
$\delta_{\rm obs}$:
\begin{equation}
  \delta_{\rm obs}
  = 
  \delta_{\rm true} \frac{F_{\rm neighbor}}{F_{\rm total}},
\end{equation}
where the total flux and the flux from the neighbor are labelled as
such.
The requirement that the eclipse is produced by stars
and that $\delta_{\rm true} < 0.5$
translates to a bound on the faintest possible blended companion
stars:
\begin{equation}
    \Delta m < - \frac{5}{2} \log_{10} \left( \frac{0.5}{\delta_{\rm obs}} \right).
\end{equation}
For \tn\ ($T=9.93$), this implies that any star invoked to explain the
transit depth must be brighter than $T=15.07$.

Typically, the shape of the transit enables this argument to be 
scaled to even more restrictive depths (e.g., CITE Vanderburg+2019, Rizzuto+2020, David+2018 1298Tau).
However, since the transits of \tn\ are grazing,
these arguments do not apply.

\subsubsection{Visual Binarity}
Figure~\ref{fig:scene} shows the scene.
In the upper panels, the pixels used to measure the background level
in the SPOC lightcurve are indicated with `\texttt{X}' hatching, and
the pixels used in the final lightcurve aperture are shown with
`\texttt{/}' hatching.

Stars brighter than $T=16$, as queried from the Gaia DR2 source
catalog, are shown with orange circles.
Given its galactic latitude of only $-6^\circ$, it is not surprising
that the field of \tn\ is crowded.
The relevant stars for a blend analysis are as follows.
\begin{itemize}
  \item \tn\ = TIC 460205581 (T=9.9). The target star.
  \item Star A = TIC 847769574 (T=14.6). $2.3$'' west. Proper motions
    and parallax imply it is comoving with \tn, though with a physical
    separation of $6.6\pm 0.1\,{\rm pc}$, it may not be bound.
  \item Star B = TIC 460205587 (T=13.1). $5.4$'' north.  This is a
    giant background star.
\end{itemize}
An additional source, TIC 847769581, is $4.9$'' from the target, but
too faint (T=18.8) to be the source of the observed transit signal.

The Gaia DR2 data for Star A seems somewhat poorly behaved.  While
Star A has $G=15.1$, and $Bp=14.9$, no $Rp$ magnitude is reported.
Correspondingly, no RUWE value is available (CITE).  The astrometric
reduced $\chi^2$ ($\chi^2 / (N-5)$, for $N$ the number of good
astrometric observations) seems rather poor, at $8.6$.  We suspect
that the photometric failure to produce an $Rp$ magnitude, as well as
the poor astrometric fit, are likely due to blending with \tn.

At the angular resolution of the TESS data, if either Star A or Star B
were eclipsing binaries, they could be the sources of the transit
signal.  Coincidence stellar sources below the resolution limit of
Gaia are also a concern (CITE). To rule out these possibilities, we
took a number of approaches.

\subsubsection{High-resolution imaging}

We imaged the system using SOAR-HRCam.
{\bf Carl: please describe}.
Figure~\ref{fig:soar}.
\citet{ziegler_soar_2020}

\subsubsection{TESS analysis}

We examined the CDIPS FFI lightcurves of the target, which are
available on MAST \citep{bouma_cluster_2019}. Three lightcurves are
available, based on photometric apertures with a radius of 1, 1.5, or
2.5 pixels.  In the raw difference-image light-curves, as well as the
PCA-detrended light-curves, dips of depth $\approx$0.35\% are visible,
and their properties do not significantly vary with aperture size.
Within $\approx20$'', the dips are consistent with originating from
the target star.

\subsubsection{Seeing-limited time-series photometry}

%FIXME:
{\bf left off here}
Co-author P{.} Evans obtained a total transit that showed that, within
1.5'', the transit comes from \tn.  We can be even more certain with a
pixel-level analysis. This 2.1 arcsec neighbor is also a mid to late M
dwarf, and bound given the Gaia parallax + proper motion. The transit
duration is rather long for the signal to come from the M dwarf,
rather than the G dwarf.  I think the state of photometric followup
could be improved (there is only one good total transit), but I think
over this month and next we can pull another few partials.

A total eclipse of Ziegler's faint companion would produce a 1.4\%
dip. A 30\% eclipse, more plausible for a NEB scenario, would produce
dips at exactly the right depth (0.4\%).

dmag = 4.7, transit depth is 0.4\%. This dmag converts to a flux
fraction of 0.0132. A 30\% eclipse on T1 could therefore nominally
explain the TESS data.  % (depth_obs = depth_true * flux fraction).

However in the aperture from this green r=2px green LC, say 90\% of the
flux from T1 is excluded. The eclipse on T1 would therefore need to be
much deeper -- in fact, over 100\%. This argument holds as long as at
least 70\% of the flux from T1 is excluded, so I think we're likely in
the clear.

@@

  \item {\it Hierarchical blends}.
    Limited by i) FEROS + Veloce RVs, and ii) speckle imaging.

  At present our data do not rule out the presence of bound companion
    stars from $\approx$10--100$\,$AU, with masses in the range of
    $\approx$0.1--0.2$\,$M$_\odot$ (Figure~\ref{fig:fpscenario}).
  Roughly 2-3\% of Sun-like stars in the solar neighborhood with
    binary companions have companions in this region of parameter space
  (assuming the Raghavan+2010 CITE log-normal binary
    separation distribution, and a uniform companion mass function.)

@@

    Archival SERC-J and AAO-SES plates are available for the \tn\
    field\footnote{\url{https://archive.stsci.edu/cgi-bin/dss_form}}.
    These plates were acquired in 1982 and 1992, respectively.
    For high proper motion stars archival imagery can be used to
    detect slowly moving background stars that might be an
    astrophysical false-positive source
    ({\it e.g.}, \citealt{huang_pimen_2018}).
    However \tn\ has only moved $\approx0.7''$ between 1982 and
    present, in comparison to the $\approx2.0''$ FWHM of the target on
    the plates.
    We therefore cannot resolve it with great confidence from
    background sources not already resolved through more modern
    imaging.

@@

\subsection{Additional Companions}

>>>>

\begin{abstract}
We validate the planetary nature of the transiting planet TOI 837.01.
We characterize the system using data from the NASA {\it Transiting
  Exoplanet Survey Satellite} (TESS), the ESA Gaia mission,
  ground-based photometry, and spectroscopy from CTIO1.5/CHIRON,
  MPG2.2/FEROS, and AAT/Veloce.
We find that TOI$\,$837 is a $T=9.9$ G-dwarf in IC~2602 (the
``Southern Pleiades''). 
The star is therefore 30$\pm$20 million years old.
The validated planet is warm ($P = 8.YY\,{\rm d}$) and likely
  Jupiter-sized. 
However, its transits are grazing: $b > 0.92$ at 3$\sigma$.
From transit photometry alone, the planetary size lies within
  0.58--3.87 $R_{\rm Jup}$, due to the degeneracy
  between the planetary size and its impact parameter.
From radial velocity monitoring, we find that its mass is less than
  2.XX $M_{\rm Jup}$ (3$\sigma$).
This mass limit requires the planetary size to be below
  $\approx$2.0$R_{\rm Jup}$.
Non-planetary explanations for the data, most notably hierarchical or
  background eclipsing binaries, are argued to be highly unlikely on
  statistical grounds.
TOI 837.01 therefore tentatively joins the ranks of adolescent
  exoplanets orbiting bright host stars. Further analysis of its mass,
  atmosphere, and obliquity could confirm its planetary nature beyond
  doubt, and are worthwhile because young planets do interesting
  things.
\end{abstract}

@@

\subsection{Ground-based Time-Series Photometry}
\label{subsec:groundphot}

@@

\tn\ has been reported as a member of \cn\ by multiple independent 
investigators \citep{Kharchenko_et_al_2013,cantatgaudin_gaia_2018}.
%FIXME FIXME finish these citations
%include: Damiani+2019, Cantat-Gaudin+2020 who explicitly listed it as
%IC2602

@@

\section{Validation of \pn}
\label{subsec:validation}

@@

@@

Many validation analyses are inherently flawed because they do
not deal with the fact that you are conditioned on having observed
many stars, and having found an eclipsing system.

Figure~\ref{fig:fpscenario} provides a visual summary of the possible
astrophysical false positive scenarios.  If the \tn\ signal is not
planetary, it must come from the near-total eclipse of a background
star (a BEB), or else a hierarchical companion of the primary star
(an HEB).  It must also be verified that to the angular resolution
possible, the transit signal originates from \tn, and not a
neighboring star (an NEB).
We proceed by describing each constraint in detail.

\subsection{Limits on Companions from the Transit Depth}

In HEB and BEB scenarios, the flux from \tn\ and the true eclipsing
binary host would blend together, reducing the ``true'' eclipse depth
$\delta_{\rm true}$ to produce the observed depth
$\delta_{\rm obs}$:
\begin{equation}
  \delta_{\rm obs}
  = 
  \delta_{\rm true} \frac{F_{\rm neighbor}}{F_{\rm total}},
\end{equation}
where the total flux and the flux from the neighbor are labelled as
such.  The requirement that the eclipse is produced by stars and that
$\delta_{\rm true} < 0.5$ translates to a bound on the faintest
possible blended companion stars:
\begin{equation}
    \Delta m < - \frac{5}{2} \log_{10} \left( \frac{0.5}{\delta_{\rm obs}} \right).
\end{equation}
For \tn\ ($T=9.93$), this implies that any stellar companion invoked
to explain the transit depth must be brighter than $T=15.07$.

Typically, the shape of the transits enables this argument to be scaled to even
more restrictive depths (e.g., CITE Seager 2003, Vanderburg+2019,
Rizzuto+2020, David+2018 1298Tau).  However, since the transits of
\tn\ are grazing, these arguments do not apply.

\subsection{Limits on Companions from Imaging}

\subsubsection{Gaia Imaging}
Figure~\ref{fig:scene} shows the scene as viewed in the DSS2 plates
and by TESS.  In the upper panels, the pixels used to measure the
background level in the SPOC lightcurve are indicated with
`\texttt{X}' hatching, and the pixels used in the final lightcurve
aperture are shown with `\texttt{/}' hatching.

Stars brighter than $T=16$, as queried from the Gaia DR2 source
catalog, are shown with orange circles.  Given its galactic latitude
of $b=-6^\circ$, it is not surprising that the field of \tn\ is
crowded.  The stars that were of immediate concern for a blend
analysis were as follows.
\begin{itemize}
  \item \tn\ = TIC 460205581 (T=9.9). The target star.
  \item Star A = TIC 847769574 (T=14.6). $2.3$'' west. Proper motions
    and parallax imply it is comoving with \tn, though with a physical
    separation of $6.6\pm 0.1\,{\rm pc}$, it may not be bound.
  \item Star B = TIC 460205587 (T=13.1). $5.4$'' north.  This is a
    giant background star.
\end{itemize}
An additional source, TIC 847769581, is $4.9$'' from the target, but
too faint (T=18.8) to be the source of the observed transit signal.

The Gaia DR2 data for Star A seems poorly behaved.  While
Star A has $G=15.1$, and $Bp=14.9$, no $Rp$ magnitude is reported.
Correspondingly, no RUWE value is available (CITE).  The astrometric
reduced $\chi^2$ ($\chi^2 / (N-5)$, for $N$ the number of good
astrometric observations) seems rather poor, at $8.6$.  We suspect
that the photometric failure to produce an $Rp$ magnitude, as well as
the poor astrometric fit, are likely due to blending with \tn.

At the angular resolution of the TESS data, if either Star A or Star B
were eclipsing binaries, they could be the sources of the transit
signal.  We ruled out both Star A and Star B as hosts of the eclipse
through a detailed analysis of the ground-based seeing-limited
photometry (\S~\ref{subsec:groundphot}).

\subsubsection{Speckle Imaging}

Additional point-sources could be present beyond the angular
resolution of the Gaia source catalog.
To limit this possibility,
we imaged the system using SOAR-HRCam, as described in
\citet{ziegler_soar_2020}.
{\bf Carl: please describe}.
The bandpass was a Johnson-Cousins,
where I is centered at 879 nm and has a width of 289 nm.

Figure~\ref{fig:soar} shows the result.
Star A (TIC 847769574) is detected as expected, and no additional
companions are found.
"""
The points on the plot are measured 5-sigma detectable contrasts. The
lines are linear smoothing fits between two regimes. In general, the
sensitivity of speckle imaging to companions rises sharply from the
diffraction limit to a ``knee'' at a separation of 0.15--0.2
arcseconds, where it then continues to slowly increase out to 1.5$''$,
beyond which the speckle patterns become de-correlated. For this
target, I suspect the measurement at 1$''$ is slightly larger than
expected due to some increase in speckle noise at that radius.
"""

We convert the contrast ratios obtained through the SOAR imaging to
constraints on the masses of putative companions following Montet+2014
(CITE and others!).  We employed the Baraffe+2003 models for
substellar mass objects and the MESA isochrones for stellar mass
objects (Choi et al XXXX).  We assumed that the system age was 35 Myr,
to ensure that companions would be at a plausible state of
contraction.

Due to the specific filters of the SOAR HRcam imager, we further
assumed that all sources had blackbody spectra. While this is clearly
false, we do not readily have access to the planetary and stellar
atmosphere models needed for the correct calculation. However, the
Baraffe+2003 and MESA models do report effective temperatures and
bolometric luminosities.  We opt to use these theoretical quantities,
combined with the measured transmission functions, to calculate the
absolute magnitudes in the HRCam bandpasses (CITE Tokovinin 2018
summary paper with the bandpasses).

\subsubsection{Archival Imaging}
    Archival SERC-J and AAO-SES plates are available for the \tn\
field\footnote{\url{https://archive.stsci.edu/cgi-bin/dss_form}}.
These plates were acquired in 1982 and 1992, respectively.
For high proper motion stars archival imagery can be used to
detect slowly moving background stars that might be an
astrophysical false-positive source
({\it e.g.}, \citealt{huang_pimen_2018}).
However \tn\ has only moved $\approx0.7''$ between 1982 and
present, in comparison to the $\approx2.0''$ FWHM of the target on
the plates.
We therefore cannot resolve it with great confidence from
background sources not already resolved through more modern
imaging.

\subsection{Limits on Companions from Radial Velocities}

The combined radial velocities from FEROS, CHIRON, and Veloce 
can be used to place limits on the presence of massive, close-in
bounds companion to \tn.

We derived the relevant limits using \texttt{radvel} \citep{fulton_radvel_2018}.
We assumed circular orbits for the massive companion.
We then fitted the RV data for the semiamplitude, period, time
of conjunction,
instrument offsets, and jitter parameters.
We placed a wide log-uniform prior on the semiamplitude
and period to ensure appropriate sampling of the relevant
spaces.
($\log K \sim \mathcal{U}(1,10^5)$, $\log P \sim \mathcal{U}(0.1, 10^{15})$).

We then converting the resulting posterior in period and semiamplitude
to minimum mass and semi-major axis assuming Kepler's third law.
The resulting $3\sigma$ limits are shown in Figure~\ref{fig:fpscenario}.

\subsection{Limits on Companions from Time-Series Photometry}

\subsubsection{TESS}
We examined the CDIPS FFI lightcurves of the target, which are
available on MAST \citep{bouma_cluster_2019}. Three lightcurves are
available, based on photometric apertures with a radius of 1, 1.5, or
2.5 pixels.  In the raw difference-image light-curves, as well as the
PCA-detrended light-curves, dips of depth $\approx$0.35\% are visible,
and their properties do not significantly vary with aperture size.
Within $\approx20$'', the dips are consistent with originating from
the target star.

\subsubsection{Custom Apertures on Ground-Based Time-Series Photometry}
% See "TOI-837 : a planet!" thread, May 2020

A total eclipse of Star A would produce a 1.4\%
dip. A 30\% eclipse, more plausible for a NEB scenario, would produce
dips of exactly the right depth (0.4\%).

To assess this possibility with the ground-based photometry, we
produced lightcurves centered on \tn\ with
circular apertures of radii ranging from 0.7$''$ to 5.1$''$.
We did not detect any statistically significant variation in the depth
of the transits.

We also created lightcurves with a custom set of $r=2.1''$ apertures
with positions ranging from Star A (2.3$''$ west of \tn) to 2.3$''$
east of \tn.  We did not detect any variation of the transit depth
along this line of lightcurves, as would be expected if Star A were
the eclipse host.  The apertures east of \tn\ are particularly
constraining, as they exclude over 90\% of the flux from Star A.
The eclipse on Star A would therefore need to be excessively deep to
produce the observed eclipse depth.
We take this to indicate that \tn\ is the source of the transit signal
to within $\approx2.0''$.

\subsubsection{Color Photometry}

HEB and, to a lesser extent, BEB scenarios, can be limited
by color photometry.

Ignoring color photometry,
the most plausible HEB scenarios for TOI 837 involve pairs of eclipsing M dwarfs.

\paragraph{Simple Example}
% SEE June 1, 2020 email in "Re: TOI-837.01" thread
As a simple example,
assume $M_2=M_3=0.2M_\odot$ in a twin HEB.
Assume PMS evolution, and convert between Mstar, delta mag, and Teff using MIST isochrones for a 50 Myr system. 
Take transmission functions (filters only; no instrument or atmosphere response) from the SVO filter profile service (CITE).
Integrate up blackbody functions over each bandpass for each star, and get estimates of the eclipse depths if there were a total eclipse (of either of the secondary or tertiary) with this set of parameters.
The apparent eclipse depths in each bandpass were:

BessellU: 5.03e-04
BessellB: 1.39e-03
BessellV: 3.31e-03
CousinsR: 5.59e-03
CousinsI: 9.69e-03
TESS: 9.18e-03

The relative depths are what are important (geometry lets them all be scaled to match the observed TESS depth of 40 ppt). The attached plot shows each bandpass, where the narrow ones are as in the order above, and the wide one is TESS.

U and B are ~10x shallower than TESS-band, because the M dwarf blackbody function turns over at ~2x longer wavelengths than the G dwarf blackbody (Wien's law).

Therefore if TOI 837 were a hierarchical eclipsing M dwarf pair, Phil's two R-band observations should have shown eclipses a factor of 1.64x shallower than TESS observed (so, 24ppt, instead of the ~40ppt observed). So, the 40 (+/- 10ish) ppt R-band depths seen in Phil's transits vs TESS-band argue for a planet, but the level of confidence could probably be improved with U or B.

\paragraph{Complicated Implementation}
TODO, probably after U or B-band obtained.

@@

  \item {\it Hierarchical blends}.
    Limited by i) FEROS + Veloce RVs, and ii) speckle imaging. 
    Multicolor photometry also limits.

  At present our data do not rule out the presence of bound companion
    stars from $\approx$10--100$\,$AU, with masses in the range of
    $\approx$0.15--0.25$\,$M$_\odot$ (Figure~\ref{fig:fpscenario}).
  Roughly 2-3\% of Sun-like stars in the solar neighborhood with
    binary companions have companions in this region of parameter space
  (assuming the Raghavan+2010 CITE log-normal binary
    separation distribution, and a uniform companion mass function.)

@@

@@

\subsection{Additional Transits}
